\bf Classification and Regression on Random Dot Product Graphs

### 2025-01-24 draft.Rmd
minor changes

It has been shown \citep{Koo_2022, rubindelanchy2017statistical} that the RDPG (as well as the generalized random dot product graph, or GRDPG, \citep{rubindelanchy2017statistical}) can describe a wide range of popular random graph models, such as the $\ER$ model, the stochastic block model (SBM) \citep{doi:10.1080/0022250X.1971.9989788}, degree corrected block model (DCBM) \citep{Karrer_2011}, and popularity adjusted block model (PABM) \citep{307cbeb9b1be48299388437423d94bf1}. 
In these examples, the (G)RPDG representation of these networks allows for inference and analysis in more familiar Euclidean space rather than on graphs, and the geometry of the configuration of latent vectors correspond to the parameters of these graph models. 
More specifically, when these network models are described as (G)RDPGs, each latent vector lies on a subspace. 
\citet{athreya2020estimation} and \citet{Passino_2020} extended this to the latent structure model (LSM) and latent structure block model (LSBM) to include latent configurations that consist of nonlinear manifolds.

In this paper, we apply these highly structured random dot product graph models to supervised learning problems in which there are multiple graphs, $G_1, ..., G_L$, each with a response $y^{(1)}, ..., y^{(L)}$. 
Each graph, represented by adjacency matrix $A^{(\ell)}$, is presumed to be drawn from a RDPG with parameter set $\theta_l$, and in turn, the response $y_l$ is drawn from a univariate distribution with parameter $\theta_l$.
Since the latent vectors of each $A^{(\ell)}$ are presumed to be distributed on a structured subset of the latent space, e.g., a manifold, we call this the multiple latent structure model (MLSM). 
Much previous work has been done on multilayer networks, such as the multilayer degree corrected block model of \citet{agterberg2022joint}, multiple random eigen graphs of \citet{8889404}, and the common subspace independent edge graphs of \citet{arroyo2020inference}. 
In each of these models, the main inference problem is to identify parameters that are shared across multiple graphs. 
Our contribution is in applying similar methods to identify parameters specific to each graph, which are in turn used as features for supervised learning. 
This further allows us to model graphs with differing vertex sets.

@@

### Connections to Related Models

@@

### Classification and Regression on the MLSM

To use the MLSM for classification or regression problems, we assign response variables $y_1, ..., y_L$ to each graph. 
Then if each response $y_\ell$ depends on $p^{(\ell)}(t)$ (the parameterization of the $\ell^{th}$ curve) or $\theta_\ell$ (the parameter of the $\ell^{th}$ distribution), or some combination of the two, there is a plausible setup for a predictive modeling task for predicting $y_\ell$ after observing $A^{(\ell)}$. 
Four such scenarios are given:

\begin{definition}[MLSM for classification 1]
\label{def:mlsm-c-1}
Let $A^{(1)}, ..., A^{(L)} \sim \MLSM(\{p^{(\ell)}\}, F, \{\theta_\ell\}_L)$, $y_1, ..., y_L \in \{1, ..., M\}$ be labels for each of the $L$ graphs of the MLSM, and each $\theta_\ell$ take on one of $M$ values $\{\phi_1, ..., \phi_M\}$ corresponding to its response, $y_\ell$, i.e., $\theta_\ell = \phi_{y_\ell}$.

We observe the adjacency matrices $A^{(1)}, ..., A^{(L)}$ and the first $r$ labels $y_1, ..., y_r$, and the task is to predict the unknown labels $y_{r+1}, ..., y_L$.

This can be described by the following generative model:

\begin{enumerate}
  \item Draw labels $y_1, ..., y_L \iid \Categorical(\{1, ..., M\}, \{\pi_1, ..., \pi_M\})$.
  \item Set each $\theta_\ell = \phi_{y_\ell}$.
  \item Draw adjacency matrices as $A^{(1)}, ..., A^{(L)} \sim \MLSM(\{p^{(\ell)}\}, F, \{\theta_\ell\}_L)$.
\end{enumerate}
\end{definition}

\begin{definition}[MLSM for classification 2]
\label{def:mlsm-c-2}
Let $A^{(1)}, ..., A^{(L)} \sim \MLSM(\{\mathcal{C}^{(\ell)}\}, F, \{\theta_\ell\}_L)$, $y_1, ..., y_L \in \{1, ..., M\}$ be labels for each of the $L$ graphs of the MLSM, and each $\mathcal{C}^{(\ell)}$ take on one of $M$ functional forms $\{p^{(1)}(t), ..., p^{(M)}(t)\}$ corresponding to its response variable, $y_\ell$, i.e., $\mathcal{C}^{(\ell)}$ is parameterized by $p^{(y_\ell)}(t)$.

We observe the adjacency matrices $A^{(1)}, ..., A^{(L)}$ and the first $r$ labels $y_1, ..., y_r$, and the task is to predict the unknown labels $y_{r+1}, ..., y_L$.

This can be described by the following generative model:

\begin{enumerate}
  \item Draw labels $y_1, ..., y_L \iid \Categorical(\{1, ..., M\}, \{\pi_1, ..., \pi_M\})$.
  \item Set each $p^{(\ell)}(t) = p^{(y_\ell)}(t)$.
  \item Draw adjacency matrices as $A^{(1)}, ..., A^{(L)} \sim \MLSM(\{p^{(\ell)}\}, F, \{\theta_\ell\}_L)$.
\end{enumerate}

@@

### Estimation of the LSM Curves

@@

\begin{algorithm}[H]
\label{alg:bezier-fit-mixture}
\DontPrintSemicolon
\SetAlgoLined
\setstretch{1.1}
\KwData{Adjacency matrix $A$, embedding dimension $d$, polynomial degree $R$, number of latent structures $K$, stopping criterion $\epsilon$.}
\KwResult{Bezier coefficients $\hat{b}$, timepoint values $\hat{t}_1, ..., \hat{t}_n$, curve memberships $\hat{z}_1, ..., \hat{z}_n$.}
Compute $\hat{X} \in \mathbb{R}^{n \times d}$, the ASE of $A$.\;
Initialize $\hat{z}_1, ..., \hat{z}_n \in \{1, ..., K\}$, the membership of each vertex to each latent curve.\;
\For {$k = 1, ..., K$} {
  Subset the rows of $\hat{X}$ by $z_i = k$ to obtain $\hat{x}_{k_1}, ..., \hat{x}_{k_{n_k}}$, where $n_k$ is the number of vertices with label $k$.\;
  Initialize timepoints $\hat{t}_{k_1}, ..., \hat{t}_{k_{n_k}}$, (e.g., via a one-dimensional Isomap embedding \citep{Tenenbaum2000-ff}).\;
}
\Repeat {the change in equation \ref{eq:bezier-ls} is less than $\epsilon$.} {
  \For {$k = 1, ..., K$} {
    Fit $\hat{b}^{(k)}$ by equation \ref{eq:bezier-ols}, using the row subsets of $T$ and $X$ such that $z_i = k$.\;
    \For {$i = 1, ..., n_k$} {
      Fit $\hat{t}_{k_i}$ by equation by finding the roots of equation \ref{eq:dbezier} and choosing the roots that minimizes equation \ref{eq:bezier-ls}.\;
    }
  }
  \For {$k = 1, ..., n$} {
    Reassign $z_i$ based on the latent curve closest to $\hat{x}_i$. 
  }
}
\caption{Procedure for estimating multiple LSM curves as a Bezier polynomial from an adjacency matrix.}
\end{algorithm}

@@

### Quasi-Maximum Likelihood Estimation for $\theta$

@@

### Classification and Regression

@@

\begin{algorithm}[H]
\label{alg:mlsm-fit}
\DontPrintSemicolon
\SetAlgoLined
\setstretch{1.1}
\KwData{Adjacency matrices $A^{(1)}, ..., A^{(L)}$, response variables $y_1, ..., y_L$, embedding dimension $d$, probability distribution family $F$, polynomial degree $R$, stopping criterion $\epsilon$.}
\KwResult{Bezier coefficients $\hat{b}$, timepoint values $\hat{t}_1, ..., \hat{t}_n$.}
\For {$\ell = \{1, ..., L\}$} {
  Compute $\hat{X}^{(\ell)} \in \mathbb{R}^{n_\ell \times d}$, the $d$-dimensional ASE of $A^{(\ell)} \in \mathbb{R}^{n_\ell \times n_\ell}$.\;
  Initialize timepoints $\hat{t}_1, ..., \hat{t}_{n_\ell}$;
  \Repeat {the change in equation \ref{eq:bezier-ls} is less than $\epsilon$.} {
    Fit $\hat{b}$ by equation \ref{eq:bezier-ols}.\;
    \For {$i = 1, ..., {n_\ell}$} {
      Fit $\hat{t}_i$ by equation by finding the roots of equation \ref{eq:dbezier} and choosing the roots that minimizes equation \ref{eq:bezier-ls}.\;
    }
  }
  Estimate $\hat{\theta}^{(\ell)}$ from $\hat{t}_1, ..., \hat{t}_{n_\ell}$ via maximum likelihood estimation.\;
}
Fit a model for predicting $y^{(\ell)}, ..., y^{(\ell)}$ using the estimated parameters $\theta^{(\ell)}, ..., \theta^{(\ell)}$ (e.g., via linear regression).\;
\caption{Procedure for fitting a classification or regression model for an MLSM.}
\end{algorithm}

@@

In these simulations, the responses $y^{(\ell)}$ are generated as linear responses of $\theta^{(\ell)}$ (in scenario 1) or $a^{(\ell)}$ and $b^{(\ell)}$ (in scenario 2). 
More specifically, in scenario 1, each $y_\ell \indep \mathcal{N}(\beta_0 + \beta_1 \theta^{(\ell)}, \sigma^2)$, where $\theta^{(\ell)}$ is the angle between two line segments in the latent space, and in scenario 2, each $y_\ell \indep \mathcal{N}(\beta_0 + \beta_1 a^{(\ell)} + \beta_2 b^{(\ell)}, \sigma^2)$, where $a^{(\ell)}$ and $b^{(\ell)}$ describe the Beta parameters of the distribution of latent vectors along a curve. 
We set $\sigma^2 = 0.25$ for both scenarios.

The simulation setup for the first scenario is as follows: For each $N = 128, 256, 512, 1024$, sample $L = 128$ graphs as:

  i. Draw $n_1, ..., n_L$ as $n_\ell \iid \Poisson(N)$.
  ii. Draw $\theta^{(1)}, ..., \theta^{(L)} \iid \Uniform(\pi / 6, \pi / 3)$.
  iii. For each $\ell = 1, ..., L$, draw latent vectors either from the line segment described by $p_1(t) = \begin{bmatrix} t & 0 \end{bmatrix}^\top$ or $p_2(t) = \begin{bmatrix} \cos(\theta^{(\ell)}) & \sin(\theta^{(\ell)}) \end{bmatrix}^\top$ with probability $1/2$ for each.
  iv. Use the $L$ sets of latent vectors to draw $L$ RDPGs with adjacency matrices $A^{(1)}, ..., A^{(L)}$.
  v. Draw each $y_\ell \indep \mathcal{N}(1 + 2 \theta^{(\ell)}, \sigma^2)$.
  vi. Apply algorithm \ref{alg:mlsm-fit} to construct a linear model connecting $y_1, ..., y_\ell$ and $A^{(1)}, ..., A^{(L)}$, using the first half as training data and reporting MSE on the last half not used for training.

For the second the simulation setup is similar. For each $N = 128, 256, 512, 1024$, sample $L = 128$ graphs as:

  i. Draw $n_1, ..., n_L$ as $n_\ell \iid \Poisson(N)$.
  ii. Draw each $(a^{(\ell)}, b^{(\ell)}) \iid \Uniform([0.5, 2] \times [0.5, 2])$.
  iii. For each $\ell = 1, ..., L$, draw latent vectors by first taking $t_1, ..., t_{n_\ell} \sim \Betadist(a^{(\ell)}, b^{(\ell)})$ then taking each latent vector as $x_i = \begin{bmatrix} t^2 & 2 t (1-t) \end{bmatrix}^\top$.
  iv. Use the $L$ sets of latent vectors to draw $L$ RDPGs with adjacency matrices $A^{(1)}, ..., A^{(L)}$.
  v. Draw each $y_\ell \indep \mathcal{N}(1 + 2 \alpha^{(\ell)} + b^{(\ell)}, \sigma^2)$.
  vi. Apply algorithm \ref{alg:mlsm-fit} to construct a linear model connecting $y_1, ..., y_\ell$ and $A^{(1)}, ..., A^{(L)}$, using the first half as training data and reporting MSE on the last half not used for training.

In figure \ref{fig:angle-regression-sim}, we compare algorithm \ref{alg:mlsm-fit} to fitting a regression model using graph transitivity and average degree of each vertex. 
Since this scenario is similar to a multilayer DCBM, we expect performance to be very similar, which is what we observe in the results. 
However, we cannot model this simulation as a multilayer DCBM since although each graph can be described as a DCBM, the graphs do not share the same vertex set. 
In figure \ref{fig:beta-regression-sim}, we see that linear regression using graph summary statistics performs better for smaller graphs, but as we draw more vertices, the ASE improves and so algorithm \ref{mlsm-fit} performs better, whereas graph summary statistics cannot improve with the number of vertices. 
Figures \ref{fig:theta-mse} and \ref{fig:beta-mse} show that the quasi-MLE estimators computed via algorithm \ref{alg:bezier-param-fit} are consistent.

>>>>

It has been shown that the RDPG (as well as the generalized random dot product graph, or GRDPG, \citep{rubindelanchy2017statistical}) can describe a wide range of popular random graph models, such as the $\ER$ model, the stochastic block model (SBM) \citep{doi:10.1080/0022250X.1971.9989788}, degree corrected block model (DCBM) \citep{Karrer_2011}, and popularity adjusted block model (PABM) \citep{307cbeb9b1be48299388437423d94bf1}. 
In this paper, we apply these highly structured random dot product graph models to supervised learning problems in which there are multiple graphs, $G_1, ..., G_L$, each with a response $y^{(1)}, ..., y^{(L)}$. 
Each graph, represented by adjacency matrix $A^{(\ell)}$, is presumed to be drawn from a RDPG with parameter set $\theta_l$, and in turn, the response $y_l$ is drawn from a univariate distribution with parameter $\theta_l$.
Since the latent vectors of each $A^{(\ell)}$ are presumed to be distributed on a structured subset of the latent space, e.g., a manifold, we call this the multiple latent structure model (MLSM). 
There exists much existing literature on the estimation for multilayer networks, such as the multilayer degree corrected block model of \citet{agterberg2022joint}, multiple random eigen graphs of \citet{8889404}, and the common subspace independent edge graphs of \citet{arroyo2020inference}. 
In each of these models, the main inference problem is to identify parameters that are shared across multiple graphs. 
Our contribution is in applying similar methods to identify parameters specific to each graph, which are in turn used as features for supervised learning. 
This further allows us to model graphs with differing vertex sets.

@@

@@

To use the MLSM for regression problems, we assign response variables $y_1, ..., y_L$ to each graph. 
Then if each response $y_\ell$ depends on $p^{(\ell)}(t)$ (the parameterization of the $\ell^{th}$ curve) or $\theta_\ell$ (the parameter of the $\ell^{th}$ distribution), or some combination of the two, there is a plausible setup for a predictive modeling task for predicting $y_\ell$ after observing $A^{(\ell)}$.

@@

@@

@@

@@

@@

\begin{algorithm}[H]
\label{alg:mlsm-fit}
\DontPrintSemicolon
\SetAlgoLined
\setstretch{1.1}
\KwData{Adjacency matrices $A^{(1)}, ..., A^{(L)}$, response variables $y_1, ..., y_L$, embedding dimension $d$, probability distribution family $F$, polynomial degree $R$, stopping criterion $\epsilon$.}
\KwResult{Bezier coefficients $\hat{b}$, timepoint values $\hat{t}_1, ..., \hat{t}_n$.}
\For {$\ell = \{1, ..., L\}$} {
  Compute $\hat{X}^{(\ell)} \in \mathbb{R}^{n_\ell \times d}$, the $d$-dimensional ASE of $A^{(\ell)} \in \mathbb{R}^{n_\ell \times n_\ell}$.\;
  Initialize timepoints $\hat{t}_1, ..., \hat{t}_{n_\ell}$;
  \Repeat {the change in equation \ref{eq:bezier-ls} is less than $\epsilon$.} {
    Fit $\hat{b}$ by equation \ref{eq:bezier-ols}.\;
    \For {$i = 1, ..., {n_\ell}$} {
      Fit $\hat{t}_i$ by equation by finding the roots of equation \ref{eq:dbezier} and choosing the roots that minimizes equation \ref{eq:bezier-ls}.\;
    }
  }
  Estimate $\hat{\theta}^{(\ell)}$ from $\hat{t}_1, ..., \hat{t}_{n_\ell}$ via maximum likelihood estimation.\;
}
Fit a model for predicting $y^{(\ell)}, ..., y^{(\ell)}$ using the estimated parameters $\theta^{(\ell)}, ..., \theta^{(\ell)}$ (e.g., via linear regression).\;
\caption{Procedure for fitting a regression model for an MLSM.}
\end{algorithm}

@@

In these simulations, the responses $y^{(\ell)}$ are generated as linear responses of $\theta^{(\ell)}$ (in scenario 1) or $a^{(\ell)}$ and $b^{(\ell)}$ (in scenario 2). 
More specifically, in scenario 1, each $y_\ell \indep \mathcal{N}(\beta_0 + \beta_1 \theta^{(\ell)}, \sigma^2)$, where $\theta^{(\ell)}$ is the angle between two line segments in the latent space, and in scenario 2, each $y_\ell \indep \mathcal{N}(\beta_0 + \beta_1 a^{(\ell)} + \beta_2 b^{(\ell)}, \sigma^2)$, where $a^{(\ell)}$ and $b^{(\ell)}$ describe the Beta parameters of the distribution of latent vectors along a curve. 
We set $\sigma^2 = 0.25$ for both scenarios. 
We also vary the expected number of vertices for each graph: For each $N = 128, 256, 512$, and $1024$, we sample $L = 128$ graphs as $n_1, ..., n_L \iid \Poisson(N)$. 
Algorithm \ref{alg:mlsm-fit} is applied to each set of $128$ graphs, then half of the estimates are used for fitting the linear regression model and the mean squared error is computed on the remainder. 
To compare our algorithm against existing methods, we extracted graph statistics as covariates for a linear regression model. 
For each $N$, we ran $50$ replicates to obtain the average and standard error of each method.

In figure \ref{fig:angle-regression-sim}, we compare algorithm \ref{alg:mlsm-fit} to fitting a regression model using graph transitivity and average degree of each vertex (scenario 1). 
Since this scenario is similar to a multilayer DCBM, we expect performance to be very similar, which is what we observe in the results. 
However, we cannot model this simulation as a multilayer DCBM since although each graph can be described as a DCBM, the graphs do not share the same vertex set. 
In figure \ref{fig:beta-regression-sim}, we see that linear regression using graph summary statistics performs better for smaller graphs, but as we draw more vertices, the ASE improves and so algorithm \ref{mlsm-fit} performs better, whereas graph summary statistics cannot improve with the number of vertices. 
Figures \ref{fig:theta-mse} and \ref{fig:beta-mse} show that the quasi-MLE estimators computed via algorithm \ref{alg:bezier-param-fit} are consistent.

### 2025-01-24 draft.tex
minor changes

It has been shown \citep{Koo_2022, rubindelanchy2017statistical} that
the RDPG (as well as the generalized random dot product graph, or GRDPG,
\citep{rubindelanchy2017statistical}) can describe a wide range of
popular random graph models, such as the \(\ER\) model, the stochastic
block model (SBM) \citep{doi:10.1080/0022250X.1971.9989788}, degree
corrected block model (DCBM) \citep{Karrer_2011}, and popularity
adjusted block model (PABM) \citep{307cbeb9b1be48299388437423d94bf1}. In
these examples, the (G)RPDG representation of these networks allows for
inference and analysis in more familiar Euclidean space rather than on
graphs, and the geometry of the configuration of latent vectors
correspond to the parameters of these graph models. More specifically,
when these network models are described as (G)RDPGs, each latent vector
lies on a subspace. \citet{athreya2020estimation} and
\citet{Passino_2020} extended this to the latent structure model (LSM)
and latent structure block model (LSBM) to include latent configurations
that consist of nonlinear manifolds.

In this paper, we apply these highly structured random dot product graph
models to supervised learning problems in which there are multiple
graphs, \(G_1, ..., G_L\), each with a response
\(y^{(1)}, ..., y^{(L)}\). Each graph, represented by adjacency matrix
\(A^{(\ell)}\), is presumed to be drawn from a RDPG with parameter set
\(\theta_l\), and in turn, the response \(y_l\) is drawn from a
univariate distribution with parameter \(\theta_l\). Since the latent
vectors of each \(A^{(\ell)}\) are presumed to be distributed on a
structured subset of the latent space, e.g., a manifold, we call this
the multiple latent structure model (MLSM). Much previous work has been
done on multilayer networks, such as the multilayer degree corrected
block model of \citet{agterberg2022joint}, multiple random eigen graphs
of \citet{8889404}, and the common subspace independent edge graphs of
\citet{arroyo2020inference}. In each of these models, the main inference
problem is to identify parameters that are shared across multiple
graphs. Our contribution is in applying similar methods to identify
parameters specific to each graph, which are in turn used as features
for supervised learning. This further allows us to model graphs with
differing vertex sets.

@@

\subsubsection{Connections to Related
Models}\label{connections-to-related-models}

@@

\subsubsection{Classification and Regression on the
MLSM}\label{classification-and-regression-on-the-mlsm}

To use the MLSM for classification or regression problems, we assign
response variables \(y_1, ..., y_L\) to each graph. Then if each
response \(y_\ell\) depends on \(p^{(\ell)}(t)\) (the parameterization
of the \(\ell^{th}\) curve) or \(\theta_\ell\) (the parameter of the
\(\ell^{th}\) distribution), or some combination of the two, there is a
plausible setup for a predictive modeling task for predicting \(y_\ell\)
after observing \(A^{(\ell)}\). Four such scenarios are given:

\begin{definition}[MLSM for classification 1]
\label{def:mlsm-c-1}
Let $A^{(1)}, ..., A^{(L)} \sim \MLSM(\{p^{(\ell)}\}, F, \{\theta_\ell\}_L)$, $y_1, ..., y_L \in \{1, ..., M\}$ be labels for each of the $L$ graphs of the MLSM, and each $\theta_\ell$ take on one of $M$ values $\{\phi_1, ..., \phi_M\}$ corresponding to its response, $y_\ell$, i.e., $\theta_\ell = \phi_{y_\ell}$.

We observe the adjacency matrices $A^{(1)}, ..., A^{(L)}$ and the first $r$ labels $y_1, ..., y_r$, and the task is to predict the unknown labels $y_{r+1}, ..., y_L$.

This can be described by the following generative model:

\begin{enumerate}
  \item Draw labels $y_1, ..., y_L \iid \Categorical(\{1, ..., M\}, \{\pi_1, ..., \pi_M\})$.
  \item Set each $\theta_\ell = \phi_{y_\ell}$.
  \item Draw adjacency matrices as $A^{(1)}, ..., A^{(L)} \sim \MLSM(\{p^{(\ell)}\}, F, \{\theta_\ell\}_L)$.
\end{enumerate}
\end{definition}

\begin{definition}[MLSM for classification 2]
\label{def:mlsm-c-2}
Let $A^{(1)}, ..., A^{(L)} \sim \MLSM(\{\mathcal{C}^{(\ell)}\}, F, \{\theta_\ell\}_L)$, $y_1, ..., y_L \in \{1, ..., M\}$ be labels for each of the $L$ graphs of the MLSM, and each $\mathcal{C}^{(\ell)}$ take on one of $M$ functional forms $\{p^{(1)}(t), ..., p^{(M)}(t)\}$ corresponding to its response variable, $y_\ell$, i.e., $\mathcal{C}^{(\ell)}$ is parameterized by $p^{(y_\ell)}(t)$.

We observe the adjacency matrices $A^{(1)}, ..., A^{(L)}$ and the first $r$ labels $y_1, ..., y_r$, and the task is to predict the unknown labels $y_{r+1}, ..., y_L$.

This can be described by the following generative model:

\begin{enumerate}
  \item Draw labels $y_1, ..., y_L \iid \Categorical(\{1, ..., M\}, \{\pi_1, ..., \pi_M\})$.
  \item Set each $p^{(\ell)}(t) = p^{(y_\ell)}(t)$.
  \item Draw adjacency matrices as $A^{(1)}, ..., A^{(L)} \sim \MLSM(\{p^{(\ell)}\}, F, \{\theta_\ell\}_L)$.
\end{enumerate}

@@

\subsubsection{Estimation of the LSM
Curves}\label{estimation-of-the-lsm-curves}

@@

\subsubsection{\texorpdfstring{Quasi-Maximum Likelihood Estimation for
\(\theta\)}{Quasi-Maximum Likelihood Estimation for \textbackslash theta}}\label{quasi-maximum-likelihood-estimation-for-theta}

@@

\subsubsection{Classification and
Regression}\label{classification-and-regression}

@@

In these simulations, the responses \(y^{(\ell)}\) are generated as
linear responses of \(\theta^{(\ell)}\) (in scenario 1) or
\(a^{(\ell)}\) and \(b^{(\ell)}\) (in scenario 2). More specifically, in
scenario 1, each
\(y_\ell \indep \mathcal{N}(\beta_0 + \beta_1 \theta^{(\ell)}, \sigma^2)\),
where \(\theta^{(\ell)}\) is the angle between two line segments in the
latent space, and in scenario 2, each
\(y_\ell \indep \mathcal{N}(\beta_0 + \beta_1 a^{(\ell)} + \beta_2 b^{(\ell)}, \sigma^2)\),
where \(a^{(\ell)}\) and \(b^{(\ell)}\) describe the Beta parameters of
the distribution of latent vectors along a curve. We set
\(\sigma^2 = 0.25\) for both scenarios.

The simulation setup for the first scenario is as follows: For each
\(N = 128, 256, 512, 1024\), sample \(L = 128\) graphs as:

\begin{enumerate}
\def\labelenumi{\roman{enumi}.}
\tightlist
\item
  Draw \(n_1, ..., n_L\) as \(n_\ell \iid \Poisson(N)\).
\item
  Draw
  \(\theta^{(1)}, ..., \theta^{(L)} \iid \Uniform(\pi / 6, \pi / 3)\).
\item
  For each \(\ell = 1, ..., L\), draw latent vectors either from the
  line segment described by
  \(p_1(t) = \begin{bmatrix} t & 0 \end{bmatrix}^\top\) or
  \(p_2(t) = \begin{bmatrix} \cos(\theta^{(\ell)}) & \sin(\theta^{(\ell)}) \end{bmatrix}^\top\)
  with probability \(1/2\) for each.
\item
  Use the \(L\) sets of latent vectors to draw \(L\) RDPGs with
  adjacency matrices \(A^{(1)}, ..., A^{(L)}\).
\item
  Draw each
  \(y_\ell \indep \mathcal{N}(1 + 2 \theta^{(\ell)}, \sigma^2)\).
\item
  Apply algorithm \ref{alg:mlsm-fit} to construct a linear model
  connecting \(y_1, ..., y_\ell\) and \(A^{(1)}, ..., A^{(L)}\), using
  the first half as training data and reporting MSE on the last half not
  used for training.
\end{enumerate}

For the second the simulation setup is similar. For each
\(N = 128, 256, 512, 1024\), sample \(L = 128\) graphs as:

\begin{enumerate}
\def\labelenumi{\roman{enumi}.}
\tightlist
\item
  Draw \(n_1, ..., n_L\) as \(n_\ell \iid \Poisson(N)\).
\item
  Draw each
  \((a^{(\ell)}, b^{(\ell)}) \iid \Uniform([0.5, 2] \times [0.5, 2])\).
\item
  For each \(\ell = 1, ..., L\), draw latent vectors by first taking
  \(t_1, ..., t_{n_\ell} \sim \Betadist(a^{(\ell)}, b^{(\ell)})\) then
  taking each latent vector as
  \(x_i = \begin{bmatrix} t^2 & 2 t (1-t) \end{bmatrix}^\top\).
\item
  Use the \(L\) sets of latent vectors to draw \(L\) RDPGs with
  adjacency matrices \(A^{(1)}, ..., A^{(L)}\).
\item
  Draw each
  \(y_\ell \indep \mathcal{N}(1 + 2 \alpha^{(\ell)} + b^{(\ell)}, \sigma^2)\).
\item
  Apply algorithm \ref{alg:mlsm-fit} to construct a linear model
  connecting \(y_1, ..., y_\ell\) and \(A^{(1)}, ..., A^{(L)}\), using
  the first half as training data and reporting MSE on the last half not
  used for training.
\end{enumerate}

In figure \ref{fig:angle-regression-sim}, we compare algorithm
\ref{alg:mlsm-fit} to fitting a regression model using graph
transitivity and average degree of each vertex. Since this scenario is
similar to a multilayer DCBM, we expect performance to be very similar,
which is what we observe in the results. However, we cannot model this
simulation as a multilayer DCBM since although each graph can be
described as a DCBM, the graphs do not share the same vertex set. In
figure \ref{fig:beta-regression-sim}, we see that linear regression
using graph summary statistics performs better for smaller graphs, but
as we draw more vertices, the ASE improves and so algorithm
\ref{mlsm-fit} performs better, whereas graph summary statistics cannot
improve with the number of vertices. Figures \ref{fig:theta-mse} and
\ref{fig:beta-mse} show that the quasi-MLE estimators computed via
algorithm \ref{alg:bezier-param-fit} are consistent.

>>>>

It has been shown that the RDPG (as well as the generalized random dot
product graph, or GRDPG, \citep{rubindelanchy2017statistical}) can
describe a wide range of popular random graph models, such as the
\(\ER\) model, the stochastic block model (SBM)
\citep{doi:10.1080/0022250X.1971.9989788}, degree corrected block model
(DCBM) \citep{Karrer_2011}, and popularity adjusted block model (PABM)
\citep{307cbeb9b1be48299388437423d94bf1}. In this paper, we apply these
highly structured random dot product graph models to supervised learning
problems in which there are multiple graphs, \(G_1, ..., G_L\), each
with a response \(y^{(1)}, ..., y^{(L)}\). Each graph, represented by
adjacency matrix \(A^{(\ell)}\), is presumed to be drawn from a RDPG
with parameter set \(\theta_l\), and in turn, the response \(y_l\) is
drawn from a univariate distribution with parameter \(\theta_l\). Since
the latent vectors of each \(A^{(\ell)}\) are presumed to be distributed
on a structured subset of the latent space, e.g., a manifold, we call
this the multiple latent structure model (MLSM). There exists much
existing literature on the estimation for multilayer networks, such as
the multilayer degree corrected block model of
\citet{agterberg2022joint}, multiple random eigen graphs of
\citet{8889404}, and the common subspace independent edge graphs of
\citet{arroyo2020inference}. In each of these models, the main inference
problem is to identify parameters that are shared across multiple
graphs. Our contribution is in applying similar methods to identify
parameters specific to each graph, which are in turn used as features
for supervised learning. This further allows us to model graphs with
differing vertex sets.

@@

@@

To use the MLSM for regression problems, we assign response variables
\(y_1, ..., y_L\) to each graph. Then if each response \(y_\ell\)
depends on \(p^{(\ell)}(t)\) (the parameterization of the \(\ell^{th}\)
curve) or \(\theta_\ell\) (the parameter of the \(\ell^{th}\)
distribution), or some combination of the two, there is a plausible
setup for a predictive modeling task for predicting \(y_\ell\) after
observing \(A^{(\ell)}\).

@@

@@

@@

@@

In these simulations, the responses \(y^{(\ell)}\) are generated as
linear responses of \(\theta^{(\ell)}\) (in scenario 1) or
\(a^{(\ell)}\) and \(b^{(\ell)}\) (in scenario 2). More specifically, in
scenario 1, each
\(y_\ell \indep \mathcal{N}(\beta_0 + \beta_1 \theta^{(\ell)}, \sigma^2)\),
where \(\theta^{(\ell)}\) is the angle between two line segments in the
latent space, and in scenario 2, each
\(y_\ell \indep \mathcal{N}(\beta_0 + \beta_1 a^{(\ell)} + \beta_2 b^{(\ell)}, \sigma^2)\),
where \(a^{(\ell)}\) and \(b^{(\ell)}\) describe the Beta parameters of
the distribution of latent vectors along a curve. We set
\(\sigma^2 = 0.25\) for both scenarios. We also vary the expected number
of vertices for each graph: For each \(N = 128, 256, 512\), and
\(1024\), we sample \(L = 128\) graphs as
\(n_1, ..., n_L \iid \Poisson(N)\). Algorithm \ref{alg:mlsm-fit} is
applied to each set of \(128\) graphs, then half of the estimates are
used for fitting the linear regression model and the mean squared error
is computed on the remainder. To compare our algorithm against existing
methods, we extracted graph statistics as covariates for a linear
regression model. For each \(N\), we ran \(50\) replicates to obtain the
average and standard error of each method.

In figure \ref{fig:angle-regression-sim}, we compare algorithm
\ref{alg:mlsm-fit} to fitting a regression model using graph
transitivity and average degree of each vertex (scenario 1). Since this
scenario is similar to a multilayer DCBM, we expect performance to be
very similar, which is what we observe in the results. However, we
cannot model this simulation as a multilayer DCBM since although each
graph can be described as a DCBM, the graphs do not share the same
vertex set. In figure \ref{fig:beta-regression-sim}, we see that linear
regression using graph summary statistics performs better for smaller
graphs, but as we draw more vertices, the ASE improves and so algorithm
\ref{mlsm-fit} performs better, whereas graph summary statistics cannot
improve with the number of vertices. Figures \ref{fig:theta-mse} and
\ref{fig:beta-mse} show that the quasi-MLE estimators computed via
algorithm \ref{alg:bezier-param-fit} are consistent.
